---
title: "The British-American Rock Divide of the 1970s"
author: "Jonas Pronk"
date: "`r Sys.Date()`"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(flexdashboard)
library(tidyverse)
library(spotifyr)
library(dplyr)
library(ggthemes)
library(compmus)
library(pals)
library(plotly)
```

```{r data import, include=FALSE}
the_beatles <- get_playlist_audio_features("", "37i9dQZF1DZ06evO2iBPiw")

led_zeppelin <- get_playlist_audio_features("", "37i9dQZF1DZ06evO1NyWWI")

queen <- get_playlist_audio_features("", "37i9dQZF1DZ06evO0ENBD2")

pink_floyd <- get_playlist_audio_features("", "37i9dQZF1DZ06evO07zaak")

the_rolling_stones <- get_playlist_audio_features("", "37i9dQZF1DZ06evO19s0CZ")

eagles <- get_playlist_audio_features("", "37i9dQZF1DZ06evO0jXJRu")

aerosmith <- get_playlist_audio_features("", "37i9dQZF1DZ06evO4x3X2w")

bon_jovi <- get_playlist_audio_features("", "37i9dQZF1DZ06evO3154GY")

the_doors <- get_playlist_audio_features("", "37i9dQZF1DZ06evO19UBIk")

lynyrd_skynyrd <- get_playlist_audio_features("", "37i9dQZF1DZ06evO2ObHwc")

uk <-
  bind_rows(
    the_beatles |> mutate(band = "The Beatles"),
    led_zeppelin |> mutate(band = "Led Zeppelin"),
    queen |> mutate(band = "Queen"),
    pink_floyd |> mutate(band = "Pink Floyd"),
    the_rolling_stones |> mutate(band = "The Rolling Stones")
  )

us <-
  bind_rows(
    eagles |> mutate(band = "Eagles"),
    aerosmith |> mutate(band = "Aerosmith"),
    bon_jovi |> mutate(band = "Bon Jovi"),
    the_doors |> mutate(band = "The Doors"),
    lynyrd_skynyrd |> mutate(band = "Lynyrd Skynyrd")
  )

corpus <-
  bind_rows(
    uk |> mutate(country = "UK"),
    us |> mutate(country = "US")
  )
```

### Rock across the pond


This is the repository that I will be using during the course Computational Musicology, in which the analysis into my corpus will be visible. My corpus is made up of popular rock music from the 70s from the UK and the USA. To compare the two regions I made a selection of five British and five American bands who were popular during this time period. This selection came down to The Beatles, The Rolling stones, Queen, Led Zeppelin and Pink Floyd for the British bands, and Eagles, Aerosmith, Bon Jovi, The Doors and Lynyrd Skynyrd for the American bands. During this course I hope to dive deeper into the effect that the geographic region had on the genre of rock. To achieve this I will be looking at elements such as acousticness, valence, energy, rhythm, and timbre.
One potential area of investigation is the musical influences that shaped the sound of British and American rock bands during this time period. For example, British bands like Led Zeppelin and The Rolling Stones were heavily influenced by blues and folk music, while American bands like the Eagles and Lynyrd Skynyrd drew on country and southern rock influences. There is also a notable difference in popularity between the two regions. Four of the five British bands are in the top 10 of all time when it comes to album sales, whereas there are no American bands from this time period present.


I choose this corpus because a lot of my music interest lies within this time period. I think that specifically rock music is an interesting genre to look at, because it peaks during this time period. To produce representative data I have chosen one specific genre, as not to compare different genres. As I have only included the most popular bands, many influential and important rock bands will be missing in this corpus.


### The wall of chromas

```{r The Wall, echo=FALSE}
Wall <-
  get_tidy_audio_analysis("7K6xMPtAjTuLPNlJMLf5bS") |>
  select(segments) |>
  unnest(segments) |>
  select(start, duration, pitches)

Wall |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(
    aes(
      x = start + duration / 2,
      width = duration,
      y = pitch_class,
      fill = value
    )
  ) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```

***

Using the data from the corpus, I've identified a few outliers. Of all the songs in the corpus there were only two songs that scored lower than 0.1 on both energy and valence, those songs were the Pink Floyd songs "Another Brick in the Wall, Pt. 1" and "One of the Few". As for American outliers, the Aerosmith song "Young Lust" scores very high on energy and low on valence, something which goes in against the overall trend.

I've decided to focus a little more on the song "Another Brick in the Wall, Pt. 1", because of the importance it has in Pink Floyds repertoire. On the left you can see an chromagram of the song. The short intensity in the G key around 50 seconds in the track seems to correspond to a instrumental part on the guitar.


### Americans love their electric guitars

```{r Acousticness, echo=FALSE}
count_group <- data.frame(user=factor(rep(1:50, 2)), 
                          count=sample(100, 100, replace=T), 
                          group=factor(rep(LETTERS[1:20], 5)))

library(RColorBrewer)
cols <- colorRampPalette(brewer.pal(9, "Set1"))
ngroups <- length(unique(count_group$group))


corpus |>
  ggplot(aes(x = acousticness, fill = band)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.1) +
  facet_wrap(~country) + scale_fill_manual(values = cols(ngroups))
```

***

What can be concluded from this plot is that American bands seem to prefer electrical guitar dominated songs over British bands. Most of the low acousticness songs can be contributed to Aerosmith and Bon Jovi, while British bands seem to have a much more even distribution over different levels of acousticness.

### Who brings the positive vibes?

```{r EnergyValence, echo=FALSE}
newtemp <- list()
for ( i in 1:nrow(corpus)) {
newtemp[[i]] <- corpus$track.artists[[i]]$name
}
CountryEnergy <- ggplot(corpus, aes(x = valence, y = energy, fill = country, color = country, text = paste(newtemp, '<br>', track.name))) + geom_point(alpha = 0.6, shape = 21, stroke = 1) + theme_tufte() + xlim(0, 1) + ylim(0, 1) + geom_smooth()
ggplotly(CountryEnergy)
```

***

What is interesting about this plot is that for most levels of valence, American songs tend to have a higher energy level. For the most positive songs however, this relationship seems to flip as the highest energy and highest valence songs are mostly British.

